---
title: "Sample size based on Clopper-Pearson Method"
author: "Chao Cheng"
date: "2023-02-21"
output: 
    html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Clopper-Pearson method is an exact method for one-sample proportional test. The power of this type of test under given sample size `n`, type-1 error `alpha`, alternative hypothesis `alternative`, underlying real proportion `p_trt` and reference proportion `p_ctrl` can be computed via binomial enumeration

```{r power_function}
power2 <- function(n, p_trt, p_ctrl, alpha, alternative = "two.sided"){
    power <- 0
    for(i in (seq(n + 1) - 1)){
        prob <- dbinom(i, size = n, prob = p_trt)
        tmp <- binom.test(i, n, p = p_ctrl, 
                          alternative = alternative, 
                          conf.level = 1 - alpha)
        power <- power + (tmp$p.value < alpha) * prob
    }
    return(power)
}
```

Details about the reject probability of Clopper-Pearson method (the exact method) can be found in pdf notes (binomial_prob.pdf) of this repo.

## Sample size

